Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-04-25 elsevier_MANUSCRIPT.qmd
quick reviewer responses

Arizona, located in the southwestern United States, spans approximately 295,253 km² and encompasses a diverse range of landscapes, elevations, and climate regimes. The state includes portions of three major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and the Central Highlands, a transitional zone between them. This heterogeneity results in substantial variation in environmental conditions across the state. The Colorado Plateau is characterized by high-elevation desert and mountain woodlands, averaging 1,936 masl (6,352 ft), with mean temperatures ranging from -6°C (20°F) to 26°C (80°F) and annual precipitation of about 580 mm (23 in). In contrast, the Basin and Range region is lower in elevation, averaging 490 masl (1,600 ft), and features a semi-arid to arid climate, with temperatures ranging from 15°C (60°F) to 43°C (110°F) and an average annual precipitation of 200 mm (8 in) [@az_climate2024]. The Central Highlands feature a mix of mountainous terrain and interspersed basins, adding to the state's topographic and hydroclimatic complexity.

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. Though winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded. The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010]. Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Our initial training dataset comprised 7,724 observations across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

@@

: Comparison of trends for BFI for all sites split by various classifications. Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record. n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend. {#tbl-trends}

@@

![SHAP value plot of features used in final model. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

@@

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negaitve pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

@@

The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability. Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics. Regions such as the Grand Canyon and Mogollon Rim exhibit notably high BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.

>>>>

Arizona, located in the southwestern United States, spans approximately 295,253 km² and encompasses a diverse range of landscapes, elevations, and climate regimes. The state includes portions of two major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and a transitional zone, the Central Highlands, between them. This heterogeneity results in substantial variation in environmental conditions across the state. The Colorado Plateau is characterized by high-elevation desert and mountain woodlands, averaging 1,936 masl (6,352 ft), with mean temperatures ranging from -6°C (20°F) to 26°C (80°F) and annual precipitation of about 580 mm (23 in). In contrast, the Basin and Range region is lower in elevation, averaging 490 masl (1,600 ft), and features a semi-arid to arid climate, with temperatures ranging from 15°C (60°F) to 43°C (110°F) and an average annual precipitation of 200 mm (8 in) [@az_climate2024]. The Central Highlands feature a mix of mountainous terrain and interspersed basins, adding to the state's topographic and hydroclimatic complexity.

@@

Arizona's hydrology varies seasonally and spatially between its physiographic regions. In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock]. While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002]. As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral]. Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack. Though winter precipitation provides only 30% of annual average precipitation, it provides the majority of water for natural reservoirs [@sheppard2002].

@@

Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study. Streamgages were selected depending on criteria to ensure the applicability of each site. Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required. Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis. Streamgages directly downstream of major regulation (e.g., reservoirs, lakes, diversions) were excluded, based on USGS annual water data reports and site metadata \[@usgs2010\]. While some flow alteration is widespread, the focus was on removing gauges with clear, immediate regulatory impacts. As such, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact. After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area). Periods of record ranged from 10 to 112 years, with a median of 28 years.

@@

Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset. The DEM of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Our initial training dataset comprised 7,724 observations (annual observations of BFI at streamgages) across 45 variables. To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values. The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees. While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model. The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5. Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.

@@

![SHAP value plot of features used in final model. The number to the right of the feature name is the mean absolute SHAP value. Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}

@@

: Comparison of trends for BFI for all sites split by various classifications. Only sites with a significant ($\alpha \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record. n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend. {#tbl-trends}

@@

@@

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

@@

The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability. Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics. Regions such as the Grand Canyon and Mogollon Rim exhibit the highest BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
